---
title: "Class 11: Structural Bioinformatics I"
author: "Kelsey Dang"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## The PDB database for biomolecular structure data

>Q1: Download a CSV file from the PDB site (accessible from “Analyze” -> “PDB Statistics”
“by Experimental Method and Molecular Type”. 

>Determine what proportion of structures are protein?

92.17%  

Download CSV files from PDB website (“Analyze” -> “PDB Statistics” >
“by Experimental Method and Molecular Type”)
```{r}
# Read CSV
data <- read.csv("Data Export Summary.csv")
data
```

Proportion of entries from each method
```{r}
round(data$Total / sum(data$Total) * 100, 2)
```

Proportion that are protein
```{r}
round(sum(data$Proteins) / sum(data$Total) * 100, 2)
```

> Q2: Type HIV in the PDB website search box on the home page and determine how many
HIV-1 protease structures are in the current PDB?


## Section 4.2

Here we will read the 1HSG PDB structure and select the protein component and write out a new **protein-only** PDB format file. We then do the same for the ligand (i.e. known drug molecule) creating a **ligand-only** PDB file.

```{r}
library(bio3d)

pdb <- read.pdb("1hsg.pdb")
pdb
#write.pdb("1hsg.pdb")
#trim.pdb()
```

Creating protein PDB
```{r}
protein <- atom.select(pdb, "protein", value=TRUE)
write.pdb(ligand, file = "1hsg_protein.pdb")
```


Creating ligand PDB
```{r}
ligand <- atom.select(pdb, "ligand", value=TRUE)
write.pdb(ligand, file = "1hsg_ligand.pdb")
```

