---
title: "Class 7 R functions and packages"
author: "Kelsey Dang"
date: "10/22/2019"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R functions revisited

Source my functions from last day

```{r}
source("http://tinyurl.com/rescale-R")
```

```{r}
rescale(1:10)
```
```{r}
rescale(c(1, 10, 5, NA, 6))
```

Notes:  
* The function warning() - keeps on running, but throws a warning  
* The function stop() - stops running your function completely  
* Adding a '!' in front of the function call outputs the opposite  
* There are different error output messages for rescale v. rescale2, rescale2 error message is more understandable  

## A new function called **both.na()**

First make some simple input where I know the answer
```{r}
x <- c(1, 2, NA, 3, NA)
y <- c(NA, 3, NA, 3, 4)
```

Looked online and found the **is.na()** function
```{r}
is.na(x)
```

and the **which()* function tells me where the TRUE values are
```{r}
which( is.na(x) )
```

Individual output for x and y
```{r}
is.na(x)
is.na(y)
```

The **AND** function, so I can figure out how many TRUE there are combined
```{r}
is.na(x) & is.na(y)
```

Taking the **sum()** of TRUE FALSE vector will tell me how many TRUE elements I have
```{r}
sum(is.na(x) & is.na(y))
```

```{r}
sum(c(TRUE, TRUE, FALSE, TRUE))
```

Now, create a function that does the same thing, but put together called **both.na()**  
```{r}
both_na <- function(x, y)
{
  sum(is.na(x) & is.na(y))
}
```

Testing our newly created function
```{r}
x <- c(NA, NA, NA)
y1 <- c( 1, NA, NA)
y2 <- c( 1, NA, NA, NA)
```

```{r}
both_na(x, y2)
```
Note: reason why the error outputs 3, even though it is out of bounds is because of **RECYCLING**  

```{r}
# Since x is shorter, it will keep recycling the x-values of the vector until it reaches all of the length of vector y
x <- c(1, NA, NA)

# This is the recycling process for x : c(1, NA, NA, 1, NA, NA)
# So this outputs 4
y3 <- c( 1, NA, NA, NA, NA, NA)
both_na(x, y3)
```

Now, use stop() function to prevent the recycling
```{r}
length(x)
length(y3)
```

Creating the new function utilizing stop()
```{r}
both_na2 <- function(x, y)
{
  if (length(x) != length(y))
  {
    stop("Make the vectors the same length!")
  }
  sum(is.na(x) & is.na(y))
}
```

Testing our new both_na2() function with using the stop() function
```{r}
#both_na2(x, y3)
```

### Practice writing function : grade()
Write a function that determines an overall grade from a vector of student HW assignment scores dropping the lowest single alignment score

```{r}
# Student 1
s1 <- c(100, 100, 100, 100, 100, 100, 100, 90)

# Student 2
s2 <- c(100, NA, 90, 90, 90, 90, 97, 80)
```

Steps:
Calculate the grade as is
```{r}
perfect <- c(100, 100, 100, 100, 100, 100, 100, 100)
score <- sum(s1) / sum(perfect)
print(score)
```

Now, take out the lowest score
```{r}
lowest <- (s1[which.min(s1)])
print(lowest)
```

Assignment sum now dropping lowest score
```{r}
sum(s1[-which.min(s1)])
```

Find average
```{r}
mean(s1[-which.min(s1)])
```

Now, run this on s2. How do we deal with NA homework?
```{r}
which.min(s2)
s2[-which.min(s2)]
mean(s2[-which.min(s2)], na.rm = TRUE)
```

Using **any()** function
```{r}
any(is.na(s2))
```

We have our working code now turn it into a first function
```{r}
grade <- function(x)
{
  if( any(is.na(x)) )
  {
    warning("Student is missing an assignment")
  }
  mean(x[-which.min(x)], na.rm = TRUE)
}
```

Grade s1 and s2 with the new function
```{r}
grade(s1)
grade(s2)
```

Now, grade the whole class
```{r}
url <- "https://tinyurl.com/gradeinput"
hw <- read.csv(url, row.names = 1)
```

Using the **apply()** function  
1 represents row, 2 represents column
```{r}
apply(hw, 1, grade)
```

## Installing a bioconduction package

The commands on the slides did not work for the 3.6.1 version of Rstudio, the following below are the commands from bioconductor.  
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("msa")  









